---
title: "test"
author: "Me"
format: pdf
editor: visual
---

```{r include=FALSE}
library(matlib)
```

Write out the SVD using `Eqn()` and `symbolicMatrix()`. In Rmd, `Eqn()` can be given an equation label. Both of these equations are numbered. (In this example, using `echo = TRUE` gives output that intersperses the code with the equation output for some reason.)

```{r eval=FALSE}
Eqn("X=U \\lambda V", label='eqn:svd')
Eqn(symbolicMatrix("u", "n", "k", lhs = ''),
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE),
    symbolicMatrix("v", "k", "p", transpose = TRUE))
```

produces:

```{r results='asis', echo=FALSE}
Eqn("X=U \\lambda V", label='eqn:svd')
Eqn(symbolicMatrix("u", "n", "k", lhs = ''),
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE),
    symbolicMatrix("v", "k", "p", transpose = TRUE))
```

This can be referenced: As seen in Equation \ref{eqn:svd} \ldots

## aligned

You can also align the equations. Note the locations of the `&` for alignment.

Show the singular value decomposition:

```{r results='asis'}
Eqn("\\mathbf{X} & = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}",
    Eqn_newline(),
    symbolicMatrix("u", "n", "k", lhs = '&'),
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE),
    symbolicMatrix("v", "k", "p", transpose = TRUE),
    align=TRUE)
```

## showEqn

`showEqn()` can also write \LaTeX, but writes out the equations. However, but the `array` environment needs to be included inside `$$ ... $$` to be evaluated in a chunk, so this chunk is not evaluated (it causes a LaTeX error).

Perhaps look at `showEqn.R` to see if this can be fixed.

```{r}
A <- matrix(paste0("a_{", outer(1:3, 1:3, FUN  = paste0), "}"), 
            nrow=3) 
b <- paste0("b_", 1:3)
x <- paste0("x", 1:3)
```

```{r showEqn0, eval=FALSE}
showEqn(A, b, vars = x, latex=TRUE)
```

It would produce:

```         
\begin{array}{lllllll}
a_{11} \cdot x_1 &+& a_{12} \cdot x_2 &+& a_{13} \cdot x_3  &=&  b_1 \\ 
a_{21} \cdot x_1 &+& a_{22} \cdot x_2 &+& a_{23} \cdot x_3  &=&  b_2 \\ 
a_{31} \cdot x_1 &+& a_{32} \cdot x_2 &+& a_{33} \cdot x_3  &=&  b_3 \\ 
\end{array}
```

Evaluating the above code in an unnumbered \LaTeX math environment via `Eqn()`:

```{r results='asis'}
showEqn(A, b, vars = x, latex=TRUE) |> Eqn(number=FALSE)
```
