---
title: "symbolicMatrix, Eqn and matrix2latex"
output: 
    pdf_document:
        extra_dependencies: ["amsmath"]
date: "`r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(matlib)

source(here::here("dev","Eqn.R"))
```



Write out the SVD using `Eqn()` and `symbolicMatrix()`. In Rmd, `Eqn()` can be given an equation label.
Both of these equations are numbered. (In this example, using `echo = TRUE` gives output that
intersperses the code with the equation output for some reason.)

```{r eval=FALSE}
Eqn(cat("X=U \\lambda V"), label='eqn:svd')
Eqn({
    symbolicMatrix("u", "n", "k", lhs = '')
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE)
    symbolicMatrix("v", "k", "p", transpose = TRUE)
})
```

produces:

```{r results='asis', echo=FALSE}
Eqn(cat("X=U \\lambda V"), label='eqn:svd')
Eqn({
    symbolicMatrix("u", "n", "k", lhs = '')
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE)
    symbolicMatrix("v", "k", "p", transpose = TRUE)
})
```

This can be referenced:  As seen in Equation \ref{eqn:svd} \ldots

## aligned

You can also align the equations. Note the locations of the `&` for alignment.

```{r results='asis'}
Eqn({
    cat("X & =U \\lambda V \\\\ \n")
    symbolicMatrix("u", "n", "k", lhs = '&')
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE)
    symbolicMatrix("v", "k", "p", transpose = TRUE)
 }, align=TRUE)
```

## matrix2latex

`matrix2latex()` can also generate symbolic equations.

```{r}
A <- matrix(paste0('a_', 1:9), 3, 3)
b <- paste0("\\beta_", 1:3)
```

```{r results='asis', echo=FALSE}
matrix2latex(cbind(A,b)) |> Eqn(number=FALSE)
```

## showEqn

`showEqn()` can also write LaTeX, but writes out the equations.
However, but the `array` environment needs to be included
inside `$$ ... $$` to be evaluated in a chunk, so this chunk is not evaluated (it causes a LaTeX error).

Perhaps look at `showEqn.R` to see if this can be fixed.

```{r showEqn0, eval=FALSE}
A <- matrix(paste0("a_{", outer(1:3, 1:3, FUN  = paste0), "}"), 
            nrow=3) 
b <- paste0("b_", 1:3)
x <- paste0("x", 1:3)
showEqn(A, b, vars = x, latex=TRUE)
```

It would produce:

```
\begin{array}{lllllll}
a_{11} \cdot x_1 &+& a_{12} \cdot x_2 &+& a_{13} \cdot x_3  &=&  b_1 \\ 
a_{21} \cdot x_1 &+& a_{22} \cdot x_2 &+& a_{23} \cdot x_3  &=&  b_2 \\ 
a_{31} \cdot x_1 &+& a_{32} \cdot x_2 &+& a_{33} \cdot x_3  &=&  b_3 \\ 
\end{array}
```

