---
title: "Eqn() and ref() test with Quarto"
engine: knitr
---

```{r}
#| include : FALSE
library(matlib)
```

Standard equation and reference

$$
f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
$$ {#eq-binom}

See @eq-binom, and [-@eq-binom]

Syntax generated with `Eqn()`. Raw output:

```{r}
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}",
    quarto=TRUE)
```

Rendered.

```{r}
#| results: 'asis'
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}",
    quarto=TRUE)
```

With equation label and number.

```{r}
#| results: 'asis'
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}",
    quarto=TRUE, label='eq-binom2')
```

## Inline equation referencing.

- Manually with Quarto syntax (works): @eq-binom and @eq-binom2
- With `matlib` using `ref(., quarto=TRUE)`: `{r} ref("eq-binom", quarto=TRUE)` and `{r} ref("eq-binom2", quarto=TRUE)`

Will also work within an equation, not that this is particularly useful.
    
```{r}
#| results: 'asis'
"@eq-binom"
```

# Auto use `quarto=TRUE` if .qmd file used

Internally, set global to TRUE if .qmd file detected. Error raised if files have the same name but different extension (`Eqn.qmd` and `Eqn.Rmd` will raise error if in same dir).


```{r}
#| results: 'asis'
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}", label = 'eq-auto')
```


See `{r} ref("eq-auto", quarto=TRUE)` for auto-detected quarto label. Of course, the @ notation still works too; e.g., see @eq-auto for details
