---
title: "Using Eqn() and ref() test with Quarto"
engine: knitr
---

```{r}
#| include : FALSE
library(matlib)
library(car)
```

Quarto has it's own syntax for creating displayed equations with equation numbers and a way to refer to them in text.
We illustrate these here.

## Simple equations
A standard equation and reference, as a displayed equation in Quarto:

$$
f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
$$ {#eq-binom}

This can be referenced as: See @eq-binom, and [-@eq-binom]

Syntax generated with `Eqn()`. Raw output:

```{r}
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}",
    quarto=TRUE)
```

Using `results: 'asis'` in the chunk renders this as:

```{r}
#| results: 'asis'
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}",
    quarto=TRUE)
```

Add an equation label and number:

```{r}
#| results: 'asis'
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}",
    quarto=TRUE, label='eq-binom2')
```

## Inline equation referencing.

- Manually with Quarto syntax (works): @eq-binom and @eq-binom2
- With `matlib` using `ref(., quarto=TRUE)`: `{r} ref("eq-binom", quarto=TRUE)` and `{r} ref("eq-binom2", quarto=TRUE)`

Will also work within an equation, not that this is particularly useful.
    
```{r}
#| results: 'asis'
"@eq-binom"
```

## Auto use `quarto=TRUE` if .qmd file used

Internally, set global to TRUE if .qmd file detected. Error raised if files have the same name but different extension (`Eqn.qmd` and `Eqn.Rmd` will raise error if in same dir).


```{r}
#| results: 'asis'
Eqn("f\\left(k\\right) = \\binom{n}{k} p^k\\left(1-p\\right)^{n-k}", label = 'eq-auto')
```


See `{r} ref("eq-auto", quarto=TRUE)` for auto-detected quarto label. Of course, the @ notation still works too; e.g., see @eq-auto for details

## Testing `latexMatrix()`

Even this simple test doesn't work:

```{r}
Eqn("\\mathbf{X}=", latexMatrix(), quarto=TRUE)
```

Rendered:

```{r}
#| results: 'asis'
Eqn("\\mathbf{X}=", latexMatrix(), quarto=TRUE)
```


## A more complex example

Here, we illustrate the decomposition of sums of squares and product (SSP) matrices in a simple one-way MANOVA design.
```{r}
data(dogfood, package = "heplots")
str(dogfood)

#fit the model
dogfood.mod <- lm(cbind(start, amount) ~ formula, data=dogfood)
```

Do the MANOVA tests and extract the SSP_H and SSP_E:

```{r}
dogfood.aov <- Anova(dogfood.mod)
SSP_H <- dogfood.aov$SSP[["formula"]]  |> print()
SSP_E <- dogfood.aov$SSPE              |> print()
SSP_T <- SSP_H + SSP_E
```

Now, show that `SSP_T = SSP_H + SSP_E` numerically. As generated, each of these matrices have `rownames()` and `colnames()`. 

```{r}
options(print.latexMatrix = list(display.labels=FALSE))
options(digits = 4)
Eqn(latexMatrix(SSP_T), "=", latexMatrix(SSP_H), "+", latexMatrix(SSP_E),
    quarto = TRUE)
```

Rendered by Quarto:

```{r}
#| results: asis
Eqn(latexMatrix(SSP_T), "=", latexMatrix(SSP_H), "+", latexMatrix(SSP_E),
    quarto = TRUE)
```

