---
title: "Notes on a `renderMatrix()` function"
output: html_document
date: "`r Sys.Date()`"
---

It would be nice to be able to render a LaTeX equation from the console for the purpose of testing an equation using `Eqn()` before putting the result into a document. Presently, I'm using the [Lagrida LaTeX Equation Editor](https://latexeditor.lagrida.com/).

The idea is to be able to do something like:

```{r eval=FALSE}
matX <- latexMatrix()
renderMatrix(matX)
```

and see the result in an RStudio `Viewer` panel. However, I note that (unless I'm mistaken)
`Eqn()` now no longer has an option to return the result other than via `cat()`.


## Ideas

1. This [SO question shows the use of `kableExtra`](https://stackoverflow.com/questions/69822328/is-it-possible-to-view-an-latex-table-in-the-viewer-pane) to render a LaTeX table

See result in a Viewer pane:
```{r kable1, eval=FALSE}
library(kableExtra)
kable(head(mtcars), booktabs = T) |>
     kable_styling(latex_options = c("striped", "scale_down")) 
```

Display latex code in console:
```{r kable2}
library(kableExtra)
kable(mtcars, format = "latex", booktabs = T) |>
     kable_styling(latex_options = c("striped", "scale_down")) 
```


For this, would need to look into the internals of `kableExtra::kable()` to see how this is done.

2. The [texPreview package](https://yonicd.github.io/texPreview/) seems to do exactly what I want.
But it seems to require `TexLive` and the `tinytex` package. I'm using `MikTeX` and am
reluctant to switch to test this.

```{r eval=FALSE}
if(!(require(texPreview))) install.packages("texPreview")
library(texPreview)

texPreview::check_requirements()
texPreview::tex_preview(matX)
```

3. A function `renderMatrix()` to: 
* Insert result of `latexMatrix()` or `Eqn()` into a document template
* Compile that with `knitr`
* Preview result with `browseURL()`

4. [LaTeX formula in Shiny panel](https://stackoverflow.com/questions/30169101/latex-formula-in-shiny-panel) shows use of a `withMathJax()` function.


