---
title: "Visualizing the SVD"
author: "Duncan Murdoch"
date: "February 2, 2015"
output: 
  html_document
---

These notes are taken from a presentation by Duncan Murdoch, titled "Recent Developments in rgl".

## The Singular Value Decomposition

For a square $n \times n$ matrix $A$, the SVD is
$$ A = U D V^T $$
where 

- $U$ and $V$ are $n \times n$ orthogonal matrices (i.e. $U^TU = V^TV = I$)
- $D$ is an $n \times n$ diagonal matrix with non-negative
entries

## Displaying a Matrix Graphically

- Matrices are representations of
linear operators on vector spaces.
- The matrix $A$ is characterized by the behaviour of $y = A x$
as we vary $x$.
- Use the `rgl` package to develop a graphical representation of 
$3 \times 3$ matrices, e.g.
$$ A = \left(\begin{array}{rrr}
1 & 0.1 & 0.1 \\
2 & 1 & 0.1 \\
0.1 & 0.1 & 0.5
\end{array}\right) $$

## Attempt 1:  display the actions on the basis vectors

```{r rgl=TRUE,fig.height=2}
A <- matrix(c(1,2,0.1, 0.1,1,0.1, 0.1,0.1,0.5), 3,3)
basis <- cbind(c(0,1,0,0,0,0), c(0,0,0,1,0,0), c(0,0,0,0,0,1))
segments3d(basis, lwd = 3)
segments3d(basis %*% t(A), col = "red", lwd = 5)
text3d(1.1*basis, texts = c("","x", "", "y", "", "z"), cex = 2)
text3d(1.1*basis %*% t(A), col = "red", 
       texts=c("", "x", "", "y", "", "z"), cex = 2)
```

## Attempt 2:  coloured ellipsoids.

```{r rgl=TRUE, echo=FALSE}
sphere <- subdivision3d(cube3d(color=rep(rainbow(6),rep(4*4^4,6))),depth=4)
sphere$vb[4,] <- apply(sphere$vb[1:3,], 2, function(x) sqrt(sum(x^2)))
mult <- function(matrix, obj) transform3d(obj, t(matrix))
mat <- matrix(1:6, 2, 3)
layout3d(rbind(mat,mat+6), height = c(3,1,3,1), sharedMouse=TRUE)
shade3d(sphere)
next3d()
text3d(0,0,0, "Identity", cex=1.5)
next3d()
next3d(reuse=FALSE)
next3d(reuse=FALSE)
shade3d(mult(A, sphere))
next3d()
text3d(0,0,0,"A", cex=1.5)

svd <- svd(A)
U <- svd$u
D <- diag(svd$d)
V <- svd$v
next3d()
shade3d(mult(U, sphere))
next3d()
text3d(0,0,0, "U", cex=1.5)
next3d()
shade3d(mult(D, sphere))
next3d()
text3d(0,0,0, "D", cex=1.5)
next3d()
shade3d(mult(V, sphere))
next3d()
text3d(0,0,0, "V", cex=1.5)
```

