---
title: "Can I Number Equations in a Vignette?"
author: An Author
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
  base_format: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Can I Number Equations in a Vignette?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(matlib)
```

HTML documents use a different notation for labels and equation references. This is the cannonical way.

```
\begin{equation} 
(\#eq:binom)
  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
\end{equation} 

The equation \@ref(eq:binom) ...
```

which gives

\begin{equation} 
(\#eq:binom)
  f\left(k\right) = \binom{n}{k} p^k\left(1-p\right)^{n-k}
\end{equation} 

The equation \@ref(eq:binom) ...

`Eqn()` now detects that an HTML document is being compiled and changes the labels to match. This means that `\@ref(eq:binom)` should be used instead of the LaTeX `\ref{eq:binom}` way.

```{r results='asis'}
symbolicMatrix(lhs = "\\mathbf{X}") |> Eqn(label="eq:X")
```

Can I reference this as: See \@ref(eq:X)


