---
title: "Properties of determinants"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Properties of determinants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
options(digits=4)
par(mar=c(5,4,1,1)+.1)
```

The following examples illustrate the basic properties of the determinant of a matrix.

### Create a 2 x 2 matrix

```{r }
A <- matrix(c(3, 1, 
              2, 4), nrow=2, byrow=TRUE)
A
det(A)
```

### 1. Interchange two rows or cols changes the sign: -> -1 * det(A)

```{r }
det(A[ 2:1, ])
det(A[, 2:1 ])
```

### 2. transpose -> det (A) unchanged

```{r }
det( t(A) )
```

### 3. multiply row * k -> k * det(A)

Note that to multiply rows by *different* constants requires a diagonal matrix on the left.

```{r }
diag(c(3, 1)) %*% A
det( diag(c(3, 1))  %*% A)
```

### 4. multiply matrix * k -> k^2 * det(A)

This is because multipling a matrix by a constant multiplies **each** row.

```{r }
det(3 * A)
3^2 * det(A)
```

### 5. det (A  B) ->  det(A) * det(B)

The determinant of a product is the product of the determinants.  The same holds for
any number of terms in a matrix product.

```{r }
B <- matrix(c(4, 2, 
             3, 5), nrow=2, byrow=TRUE)
B
det(A %*% B)
det(A) * det(B)
```

### 6. proportional rows or columns -> `det() == 0`

Here we just add an additional copy of column 1 of a matrix, so `C[,3] == C[,1]`.
The determinant is 0 because the columns are linearly dependent.

```{r }
C <- matrix(c(1, 5, 
             2, 6,
             4, 4), nrow=3, byrow=TRUE)
C <- cbind(C, C[,1])
C
det(C)
```

### 7. Add multiple of one row to another -> det unchanged

This is the principle behind elementary row operations.

```{r }
A[2,] <- A[2,] - 2*A[1,]
det(A)
```

