---
title: "Generalized inverse"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized inverse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.height = 5,
  fig.width = 5
)
options(digits=4)
par(mar=c(5,4,1,1)+.1)
```

In matrix algebra, the inverse of a matrix is defined only for *square* matrices,
and if a matrix is *singular*, it does not have an inverse.

The **generalized inverse** (or *pseudoinverse*)
is an extension of the idea of a matrix inverse,
which has some but not all the properties of an ordinary inverse.

A common use of the pseudoinverse is to compute a 'best fit' (least squares)
solution to a system of linear equations that lacks a unique solution.

```{r }
library(matlib)
```

 Construct a square, *singular* matrix [See: Timm, EX. 1.7.3]

```{r }
A <-matrix(c(4, 4, -2,
             4, 4, -2,
            -2, -2, 10), nrow=3, ncol=3, byrow=TRUE)
det(A)
```

The rank is 2, so `inv(A)` won't work

```{r }
R(A)
```

 In the echelon form, this rank deficiency appears as the final row of zeros

```{r }
echelon(A)
```

`inv()` will throw an error

```{r }
try(inv(A))
```

 A **generalized inverse** does exist for any  matrix,
 but unlike the ordinary inverse, the generalized inverse is not unique, in the
 sense that there are various ways of defining a generalized inverse with
 various inverse-like properties.  The function `matlib::Ginv()` calculates
 a *Moore-Penrose* generalized inverse.

```{r }
(AI <- Ginv(A))
```

We can also view this as fractions:

```{r }
Ginv(A, fractions=TRUE)
```

### Properties of generalized inverse (Moore-Penrose inverse)

The generalized inverse is defined as the matrix $A^-$ such that

* $A * A^- * A = A$ and
* $A^- * A * A^- = A^-$

```{r }
A %*% AI %*% A
AI %*% A %*% AI
```

In addition, both $A * A^-$ and $A^- * A$ are symmetric, but
neither product gives an identity matrix, `A %*% AI != AI %*% A != I`


```{r }
zapsmall(A %*% AI)
zapsmall(AI %*% A)
```

## Rectangular matrices

For a *rectangular matrix*, $A^- = (A^{T} A)^{-1} A^{T}$
is the generalized inverse of $A$
if $(A^{T} A)^-$  is the ginv of $(A^{T} A)$ [See: Timm: EX 1.6.11]

```{r }
A <- cbind( 1, matrix(c(1, 0, 1, 0, 0, 1, 0, 1), nrow=4, byrow=TRUE))
A
```

This $4 \times 3$ matrix is not of full rank, because columns 2 and 3 sum to column 1.

```{r }
R(A)

(AA <- t(A) %*% A)
(AAI <- Ginv(AA))
```

The generalized inverse of $A$ is $(A^{T} A)^- A^{T}$,  `AAI * t(A)`

```{r }
AI <- AAI  %*%  t(A)
```

Show that it is a generalized inverse:

```{r }
A %*% AI %*% A
AI %*% A %*% AI
```

