---
title: "Inverse of a matrix"
author: "Michael Friendly"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inverse of a matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
options(digits=4)
par(mar=c(5,4,1,1)+.1)
```

The following examples illustrate the basic properties of the inverse of a matrix.

### Load the `matlib` package

This defines: `inv()`, `Inverse()`;  the standard function for matrix inverse is `solve()`

```{r }
library(matlib)
```

### Create a 3 x 3 matrix

```{r }
    A <- matrix( c(5, 1, 0,
                   3,-1, 2,
                   4, 0,-1), nrow=3, byrow=TRUE)
   det(A)
```

### 1.  det(A) != 0, so inverse exists

```{r }
   (AI  <- inv(A))
```

### 2. Definition of the inverse: $A^{-1} A = A A^{-1} = I$ or `AI * A = diag(nrow(A))`

NB:  Sometimes you will get very tiny off-diagonal values (like `1.341e-13`). The
function `zapsmall()` will round those to 0.

```{r }
   AI %*% A
```

### 3. Inverse is **reflexive**: inv(inv(A)) = A

```{r }
   inv(AI)
```

### 4. inv(A) is **symmetric** if and only if A is symmetric

```{r }
   inv( t(A) )
   isSymmetric(A)
   isSymmetric( inv( t(A) ) )
```

Here is a symmetric case

```{r }
   B <- matrix( c(4, 2, 2,
                  2, 3, 1,
                  2, 1, 3), nrow=3, byrow=TRUE)
   inv(B)
   inv( t(B) )
   isSymmetric(B)
   isSymmetric( inv( t(B) ) )
   all.equal( inv(B), inv( t(B) ) )
```

## More properties of matrix inverse
### 1. inverse of diagonal matrix = diag( 1/ diagonal)

```{r }
   D <- diag(c(1, 2, 4))
   inv(D)
   diag(1 / c(1, 2, 4))
```

### 2. INVERSE of an INVERSE: inv(inv(A)) = A

```{r }
   A <- matrix(c(1, 2, 3,  2, 3, 0,  0, 1, 2), nrow=3, byrow=TRUE)
   AI <- inv(A)
   inv(AI)
```

### 3. inverse of a **transpose**: inv(t(A)) = t(inv(A))

```{r }
   inv( t(A) )
   t( inv(A) )
```

### 4. inverse of a scalar * matrix:  inv( k*A ) = (1/k) * inv(A)

```{r }
   inv(5 * A)
   (1/5) * inv(A)
```

### 5. inverse of a matrix product: inv(A * B) = inv(B) * inv(A)

```{r }
   B <- matrix(c(1, 2, 3, 1, 3, 2, 2, 4, 1), nrow=3, byrow=TRUE)
   C <- B[, 3:1]
   A %*%  B
   inv(A %*%  B)

   inv(B) %*%  inv(A)
```

This extends to any number of terms: the inverse of a product is the product
of the inverses in reverse order.

```{r }
   (ABC <- A %*% B %*% C)
   inv(A %*% B %*% C)
   inv(C) %*% inv(B) %*% inv(A)
   inv(ABC)
```

### 6. $\det (A^{-1}) = 1 / \det(A) = [\det(A)]^{-1}$

The determinant of an inverse is the inverse of the determinant

```{r }
  det(AI)
  1 / det(A)
```

