---
title: "LaTeX Equations with symbolicMatrix, Eqn and matrix2latex"
author: Phil Chalmers, Michael Friendly
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: false
    toc: true
vignette: >
  %\VignetteIndexEntry{LaTeX Equations with symbolicMatrix, Eqn and matrix2latex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
options(digits=4)
par(mar=c(5,4,1,1)+.1)
```


<!-- --- -->
<!-- title: "LaTeX Equations with symbolicMatrix, Eqn and matrix2latex" -->
<!-- output:  -->
<!--     pdf_document: -->
<!--         extra_dependencies: ["amsmath"] -->
<!-- author: Phil Chalmers, Michael Friendly -->
<!-- date: "`r format(Sys.time(), '%d %B, %Y')`" -->
<!-- --- -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(matlib)
```


LaTeX is the de facto standard for communication and publication in scientific documents and it is very easy to typeset mathematical expressions
like Pythagoras' theorem, $a^2 + b^2 = c^2$ (using `a^2 + b^2 = c^2`) once you learn
the notation. Writing equations with arrays, matrices and vectors is
somewhat more challenging. Many people rely on interactive LaTeX editors like [Overleaf](https://www.overleaf.com/), [MathType](https://www.wiris.com/en/mathtype/), or online
versions that provide a menu-driven interface with fill-in templates for
matrices.

There are already some tools available in R for producing LaTeX output for
tables (`xtable::xtable()`), R objects (`Hmisc::latex()`)
and statistical models (`equatiomatic::extract_eq()`).

The `matlib` package extends these, providing a collection of functions that simplify using LaTeX notation for matrices, vectors and equations in documentation and in writing:

* `symbolicMatrix()`: Constructs the LaTeX code for a symbolic matrix, whose elements are a symbol, with row and column subscripts.
* `matrix2latex()`: Constructs the LaTeX code for a symbolic matrix, whose elements are a symbol, with row and column subscripts
* `showEqn()`: Shows what matrices $\mathbf{A}, \mathbf{b}$ look like as the system of linear equations, $\mathbf{A x} = \mathbf{b}$, but written out as a set of equations.
* `Eqn()`: A wrapper to produce LaTeX expressions that can be copied/pasted into documents or
used directly in `.Rmd` or `.qmd` documents to compile to equations.

When used directly in R, they produce their output to the console (using `cat()`).
In a `.Rmd` or `.qmd` document, use the chunk options: `results='asis', echo=FALSE`
so that `knitr` just outputs the text of the equations to the document.
The rendering of the equations is handled by [`pandoc`](https://pandoc.org/).

## Using `symbolicMatrix()` and `Eqn()`

`symbolicMatrix()` constructs the LaTeX code for a symbolic matrix, whose elements are a symbol, with row and column subscripts. For example, by default (with no arguments) it produces the expression for a matrix whose elements
are $x_{ij}$, for $i = 1, 2, \cdots, n; j = 1, 2, \cdots,m$ in a LaTeX `\begin{pmatrix} ... \end{pmatrix}`
environment that looks like this:

```
\begin{pmatrix} 
  x_{11} & x_{12} & \cdots & x_{1m} \\ 
  x_{21} & x_{22} & \cdots & x_{2m} \\ 
  \vdots & \vdots &        & \vdots \\ 
  x_{n1} & x_{n2} & \cdots & x_{nm} \\ 
\end{pmatrix}
```

The code above appears in the console. To render this as a matrix in a document, this must
be wrapped in a display math environment, which is provided by `Eqn()` and used in a code
chunk with the `results = 'asis'` option, giving:

```{r eqn1,results='asis'}
symbolicMatrix() |> Eqn(number = FALSE)
```

For chunk output in a document, you will get a LaTeX error, "missing $ inserted" if you forget
to use `Eqn()` or otherwise fail to make a math environment.

Some other examples:

* A $3 \times 3$ identity matrix with square brackets, specified as an equation with a left-hand side $\mathbf{I}_3$. The first argument to
`symbolicMatrix()` can be any numeric matrix.

```{r eqn2,results='asis'}
symbolicMatrix(diag(3), lhs = "\\mathbf{I}_3", 
               matrix="bmatrix") |> 
  Eqn(number = FALSE)
```

* A row or column vector; the first argument to `symbolicMatrix()` must be a **matrix**, so wrap an R vector in `matrix()`, supplying `nrow=1` (or `ncol = 1`):

```{r eqn3,results='asis'}
symbolicMatrix(matrix(LETTERS[1:4], nrow=1)) |> Eqn(number = FALSE)
```

```{r eqn4,results='asis'}
symbolicMatrix(matrix(letters[1:3], ncol=1)) |> Eqn(number = FALSE)
```

### The SVD
As a more complicated example,
here we write out the LaTeX equation for the singular value decomposition (SVD) of a 
general $n \times p$ matrix $\mathbf{X}$
using `Eqn()` and `symbolicMatrix()`. In Rmd markup, `Eqn()` can be given an equation label.
Two calls to `Eqn()` produce separate equations in the output.

Both of these equations are numbered (by default).
(`Eqn()` uses the LaTeX `equation` environment, `\begin{equation} ... \end{equation}`,
or `equation*` if the equation is not numbered (`number = FALSE`)).
The two calls to `Eqn()` are rendered as separate equations, center aligned.

<!-- (In this example, using `echo = TRUE` gives output that -->
<!-- intersperses the code with the equation output for some reason.) -->

```{r eval=FALSE}
Eqn("\\mathbf{X} = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}^\\top", label='eqn:svd')
Eqn(symbolicMatrix("u", "n", "k", lhs = ''),
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE),
    symbolicMatrix("v", "k", "p", transpose = TRUE))
```

This produces the two numbered equations:

```{r eqn-svd,results='asis', echo=FALSE}
Eqn("\\mathbf{X} = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}^\\top", label='eqn:svd')
Eqn(symbolicMatrix("u", "n", "k", lhs = ''),
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE),
    symbolicMatrix("v", "k", "p", transpose = TRUE))
```

The matrix names in Equation (@\ref{eqn:svd}) are printed in a **boldface** math font, typically used for matrices
and vectors. Note that when using
LaTeX code in R expressions each backslash (`\`) must be doubled (`\\`) in R because `\` is the
escape character.

Note that the first equation can be referenced because it was labeled:  "As seen in Equation (@\ref{eqn:svd}) \ldots ".
(In Quarto, equation labels must be of the form `#eq-label` and equation references are of the form `@eq-label`)

### Systems of equations

As another example, the chunk below shows a system of equations $\mathbf{A} \mathbf{x} = \mathbf{b}$
written out using symbolic matrices. Note the use of `cat()` to insert arbitrary LaTeX into
the stream

```{r eqn-Axb,symbMat, results='asis'}
Eqn(symbolicMatrix("a", nrow = "m", ncol="n", matrix="bmatrix"),
    symbolicMatrix("x", nrow = "n", ncol=1),
    cat("\\quad=\\quad"),
    symbolicMatrix("b", nrow = "m", ncol=1))
```


## aligned environment

You can also align separate equations relative to some symbol like an `=` sign to show separate
steps of re-expression, using the option `Eqn(..., align=TRUE)`.

Show the singular value decomposition again, but now as two separate equations aligned after the `=`
sign. Note the locations of the `&` operator for alignment, specified as the left-hand side (`lhs`)
of the second equation.


```{r eqn-align,results='asis'}
Eqn("\\mathbf{X} & = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}^\\top",
    Eqn_newline(),
    symbolicMatrix("u", "n", "k", lhs = '&'),
    symbolicMatrix("\\lambda", "k", "k", diag=TRUE),
    symbolicMatrix("v", "k", "p", transpose = TRUE),
    align=TRUE)
```

Note that in this example, there are three calls to `symbolicMatrix()`, wrapped inside `Eqn()`.
`Eqn_newline()` emits a newline (`\\`) between equations.


## matrix2latex

The `matrix2latex()` function can also generate symbolic equations from numeric or character matrices.
For numeric matrices, it can round the values or show results as fractions.

```{r results='asis'}
A <- matrix(1:12, nrow=3, ncol=4, byrow = TRUE) / 6
matrix2latex(A, fractions = TRUE, brackets = "b") |> Eqn(number = FALSE)
```


Say we want to show the matrix $[\mathbf{A} | \mathbf{b}]$ involved in the system of equations
$\mathbf{A} \mathbf{x} = \mathbf{b}$.
Create these as a character matrix and vector:

```{r}
A <- matrix(paste0('a_', 1:9), 3, 3, byrow = TRUE) |> print()
b <- paste0("\\beta_", 1:3) |> print()
```

Then use `matrix2latex()` on `cbind(A,b)` and pipe the result of `matrix2latex()` to `Eqn()`:

```{r results='asis'}
matrix2latex(cbind(A,b)) |> Eqn(number=FALSE)
```

All the R tricks for creating and modifying matrices can be used in this way.

## showEqn

`showEqn()` is designed to show a system of linear equations, $\mathbf{A x} = \mathbf{b}$, but written out as a set of equations individually. With the option `latex = TRUE` it writes these out in LaTeX form.

<!-- `showEqn()` can also write LaTeX, but writes out the equations individually. -->
<!-- However, but the `array` environment needs to be included -->
<!-- inside `$$ ... $$` to be evaluated in a chunk, so this chunk is not evaluated (it causes a LaTeX error). -->

Here, we create a character matrix containing
the elements of a $3 \times 3$ matrix `A`, whose elements are of the form
`a_{ij}` and two character vectors, `b_i` and `x_i`. 
```{r}
A <- matrix(paste0("a_{", outer(1:3, 1:3, FUN  = paste0), "}"), 
            nrow=3) |> print()

b <- paste0("b_", 1:3)
x <- paste0("x", 1:3)
```

`showEqn(..., latex = TRUE)` produces the three equations in a single `\begin{array} ... \begin{array}`
environment.

```{r showEqn0, eval=FALSE}
showEqn(A, b, vars = x, latex=TRUE)
```

If this line was run in an R console, it would produce:

```
\begin{array}{lllllll}
a_{11} \cdot x_1 &+& a_{12} \cdot x_2 &+& a_{13} \cdot x_3  &=&  b_1 \\ 
a_{21} \cdot x_1 &+& a_{22} \cdot x_2 &+& a_{23} \cdot x_3  &=&  b_2 \\ 
a_{31} \cdot x_1 &+& a_{32} \cdot x_2 &+& a_{33} \cdot x_3  &=&  b_3 \\ 
\end{array}
```

Evaluating the above code in an unnumbered LaTeX math environment via `Eqn()` gives the
desired result:

```{r results='asis'}
showEqn(A, b, vars = x, latex=TRUE) |> Eqn(number=FALSE)
```

