---
title: "LaTeX Equations with latexMatrix, Eqn and matrix2latex"
author: Phil Chalmers, John Fox, Michael Friendly
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    base_format: rmarkdown::html_vignette
    number_sections: false
    toc: true
  pdf_document:
    toc: true
    keep_tex: true
bibliography: "references.bib"
vignette: >
  %\VignetteIndexEntry{LaTeX Equations with latexMatrix, Eqn and matrix2latex}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
options(digits=4)
par(mar=c(5,4,1,1)+.1)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(matlib)
```


LaTeX is the de facto standard for communication and publication in scientific documents and it is very easy to typeset mathematical expressions
like Pythagoras' theorem, $a^2 + b^2 = c^2$ (using `a^2 + b^2 = c^2`) once you learn
the notation. With a bit of practice, the PDF of the normal distribution can be written as

```
f(x) = \frac{1}{\sigma\sqrt{2\pi}} 
  \exp\left[ -\left(\frac{x-\mu}{2\sigma}\right)^{\!2}\,\right]
```
which renders as

$$
f(x) = \frac{1}{\sigma\sqrt{2\pi}} 
  \exp\left[ -\left(\frac{x-\mu}{2\sigma}\right)^{\!2}\,\right]
$$

Writing equations with arrays, matrices and vectors is
somewhat more challenging. Many people rely on interactive LaTeX editors like [Overleaf](https://www.overleaf.com/), [MathType](https://www.wiris.com/en/mathtype/), or online
versions like [Lagrida LaTeX Equation Editor](https://latexeditor.lagrida.com/)
that provide a menu-driven interface with fill-in templates for
matrices.

There are already some tools available in R for producing LaTeX output:

* tables (e.g, `xtable::xtable()`, `tables::toLatex()`), 
* R objects (`Hmisc::latex()`),
* statistical models (`equatiomatic::extract_eq()`),
* `knitr::knit_print()` can be used to define [methods for printing objects of class `"matrix"`](https://stackoverflow.com/questions/45591286/for-r-markdown-how-do-i-display-a-matrix-from-r-variable). 
* The [`mathpix`](https://cran.r-project.org/package=mathpix) package
can take an image of a an equation or formula and produce the LaTeX code which should generate that image.
* The [`texPreview`](https://cran.r-project.org/package=texPreview) package
compiles snippets of LaTeX directly into images from the R console to view in the RStudio viewer pane, Shiny apps and RMarkdown documents.

See [Tools for making latex tables in R](https://stackoverflow.com/questions/5465314/tools-for-making-latex-tables-in-r)
for a more comprehensive list


The `matlib` package extends these, providing a collection of functions that simplify using LaTeX notation for matrices, vectors and equations in documentation and in writing:

* `latexMatrix()`: Constructs the LaTeX code for a symbolic matrix, whose elements are a symbol, with row and column subscripts. `latexMatrix()` also supports matrices with numeric elements, and the objects it produces may be used in various kinds of matrix computations, both symbolic and numeric.
* `matrix2latex()`: Constructs the LaTeX code for a symbolic matrix, whose elements are a symbol, with row and column subscripts
* `showEqn()`: Shows what matrices $\mathbf{A}, \mathbf{b}$ look like as the system of linear equations, $\mathbf{A x} = \mathbf{b}$, but written out as a set of equations.
* `Eqn()`: A wrapper to produce LaTeX expressions or equations that can be copied/pasted into documents or used directly in `.Rmd` or `.qmd` documents to compile to equations.

When used directly in R, these functions produce their output to the console (using `cat()`).
In a `.Rmd` or `.qmd` document, use the chunk options: `results='asis', echo=FALSE`
so that `knitr` just outputs the text of the equations to the document.
The rendering of the equations is handled by [`pandoc`](https://pandoc.org/) for standard
Rmarkdown or Quarto documents.

**Note**: There are several different engines for rendering mathematics in HTML documents for the Web:
[`mathml`](https://developer.mozilla.org/en-US/docs/Web/MathML),
[`katex`](https://katex.org/), and
[`mathjax`](https://www.mathjax.org/) and others, all of which can be made to work with
`pandoc`.
The features we describe below work in standard Rmarkdown or Quarto documents.
However, some more advanced
features (horizontal and vertical lines for partitioned matrices)
require `katex`
to work with [`pkgdown`](https://pkgdown.r-lib.org/).
Equation numbers and cross-references to them still do not work in `pkgdown.
See the discussion in [this pkgdown issue](https://github.com/r-lib/pkgdown/issues/2739), 



## Using `latexMatrix()` and `Eqn()`

`latexMatrix()` constructs the LaTeX code for a symbolic matrix, whose elements are a symbol, with row and column subscripts. For example, by default (with no arguments) it produces the expression for a matrix whose elements
are $x_{ij}$, for $i = 1, 2, \cdots, n; j = 1, 2, \cdots,m$ in a LaTeX `\begin{pmatrix} ... \end{pmatrix}`
environment that looks like this:

```
\begin{pmatrix} 
  x_{11} & x_{12} & \cdots & x_{1m} \\ 
  x_{21} & x_{22} & \cdots & x_{2m} \\ 
  \vdots & \vdots &        & \vdots \\ 
  x_{n1} & x_{n2} & \cdots & x_{nm} \\ 
\end{pmatrix}
```

The code above appears in the console. To render this as a matrix in a document, this must
be wrapped in a display math environment, which is provided by `Eqn()` and used in a code
chunk with the `results = 'asis'` option, giving:

```{r eqn1,results='asis'}
latexMatrix() |> Eqn()
```

For chunk output in a document, you will get a LaTeX error, "missing $ inserted" if you forget
to use `Eqn()` or otherwise fail to make the LaTeX appear inside a math environment.

Some other examples:

* A $3 \times 3$ identity matrix with square brackets, specified as an equation with a left-hand side $\mathbf{I}_3$. The first argument to
`latexMatrix()` can be any numeric matrix.

```{r eqn2,results='asis'}
Eqn("\\mathbf{I}_3 =", latexMatrix(diag(3), matrix="bmatrix"))
```

* A row or column vector; the first argument to `latexMatrix()` must be a **matrix**, so wrap an R vector in `matrix()`, supplying `nrow=1` (or `ncol = 1`):

```{r eqn3,results='asis'}
latexMatrix(matrix(LETTERS[1:4], nrow=1), matrix="Bmatrix") |> Eqn()
```

```{r eqn4,results='asis'}
latexMatrix(matrix(letters[1:3], ncol=1), matrix = "vmatrix") |> Eqn()
```

As the above examples illustrate, the matrix delimiters can be varied using the `matrix` argument.

A wide variety of options are available for the matrix element symbols, fonts, subscripts and decorations:

```{r eqn5,results='asis'}
latexMatrix("\\mathbb{q}", 3, 3, 
            matrix = "bmatrix",
            zero.based = c(TRUE, FALSE), 
            comma=TRUE, 
            exponent="-1") |>
  Eqn()
```


### The SVD
As a more complicated example,
here we write out the LaTeX equation for the [singular value decomposition](https://en.wikipedia.org/wiki/Singular_value_decomposition) (SVD) of a 
general $n \times p$ matrix $\mathbf{X}$
using `Eqn()` and `latexMatrix()`. In Rmd markup, `Eqn()` can be given an equation 
**label** (using the `label` argument), which will both label and number the equations.

Two calls to `Eqn()` produce separate equations in the output.
Both of these equations are numbered.
(`Eqn()` uses the LaTeX `equation` environment, `\begin{equation} ... \end{equation}`,
or `equation*` if the equation does not include a `label`).
The two calls to `Eqn()` are rendered as separate equations, center aligned.

<!-- (In this example, using `echo = TRUE` gives output that -->
<!-- intersperses the code with the equation output for some reason.) -->

```{r eval=FALSE}
Eqn("\\mathbf{X} = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}^\\top", label='eq:svd')
Eqn(latexMatrix("u", "n", "k", lhs = ''),
    latexMatrix("\\lambda", "k", "k", diag=TRUE),
    latexMatrix("v", "k", "p", transpose = TRUE), label='eq:svdmats')
```

This produces the two numbered equations:[^eqn-pkgdown]

[^eqn-pkgdown]: At present equation numbers don't work in vignettes rendered as
articles by `pkgdown`.

```{r eqn-svd,results='asis', echo=FALSE}
Eqn("\\mathbf{X} = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}^\\top", label='eq:svd')
Eqn(latexMatrix("u", "n", "k"),
    latexMatrix("\\lambda", "k", "k", diag=TRUE),
    latexMatrix("v", "k", "p", transpose = TRUE), label='eq:svdmats')
```

The matrix names in Equation `r ref("eq:svd")` are printed in a **boldface** math font
(`\mathbf{}`), typically used for matrices
and vectors. Note that when using
LaTeX code in R expressions each backslash (`\`) must be doubled (`\\`) in R because `\` is the
escape character.

Note that the first equation can be referenced because it was labeled:  "As seen in Equation `r ref("eq:svd")` \ldots ". References to equations can entered in text using an inline call to
`ref()`, e.g, `` `r knitr::inline_expr('ref("eq:svd")')` ``
(In Quarto, equation labels must be of the form `#eq-label` and equation references are of the form `@eq-label`)

### Systems of equations

As another example, the chunk below shows a system of equations $\mathbf{A} \mathbf{x} = \mathbf{b}$
written out using symbolic matrices. 

```{r eqn-Axb,symbMat, results='asis'}
Eqn(latexMatrix("a", nrow = "m", ncol="n", matrix="bmatrix"),
    latexMatrix("x", nrow = "n", ncol=1),
    Eqn_hspace(mid='='),
    latexMatrix("b", nrow = "m", ncol=1))
```

Extra symmetric white space is added via `Eqn_hspace()`, which can also be used for standard spacing such as `\quad` (default size), `\,`, `'1cm'` for `\hspace{}`, etc.

Section [showEqn](#showEqn) describes another way to display systems of equations.

### `aligned` environment

You can also align separate equations relative to some symbol like an `=` sign to show separate
steps of re-expression, using the option `Eqn(..., align=TRUE)`. Alignment points are marked by
`&` in LaTeX.

Show the singular value decomposition again, but now as two separate equations aligned after the `=`
sign. Note the locations of the `&` operator for alignment, specified as the left-hand side (`lhs`)
of the second equation.


```{r eqn-align,results='asis'}
Eqn("\\mathbf{X} & = \\mathbf{U} \\mathbf{\\Lambda} \\mathbf{V}^\\top",
    Eqn_newline(),
    ' & =',
    latexMatrix("u", "n", "k"),
    latexMatrix("\\lambda", "k", "k", diag=TRUE),
    latexMatrix("v", "k", "p", transpose = TRUE),
    align=TRUE)
```

Note that in this example, there are three calls to `latexMatrix()`, wrapped inside `Eqn()`.
`Eqn_newline()` emits a newline (`\\`) between equations.


## Computing with `"latexMatrix"` objects {#computing}

Objects returned by `latexMatrix()` that have **definite** (i.e., numeric) dimensions--- for example, a 3-by-2 matrix as opposed to an "n"-by-"m" matrix---may be employed in a variety of standard symbolic and numeric matrix computations. They provide some reasonable means to compose 
meaningful matrix equations in LaTeX far easier than doing this manually, matrix by matrix.

These computations include:

* the R arithmetic operators `+` (matrix addition), `-` (matrix subtraction and negation), `*` (product of a scalar and a matrix), 
* `^` (raise to a power), and 
* `%*%` (matrix multiplication), and the functions `t()` (transpose), `determinant()`, and `solve()` (matrix inverse), 
* `%O%` (kronecker product),
* Selecting sub-matrices using indexing `X[rows, cols]` and binding rows/columns with `rbind()`, `cbind()`

There are also function equivalents of the operators that are more flexible via optional arguments.  For example, using the operator `A %*% B` multiplies the two matrices `A` and `B`, returning a symbolic result. The corresponding function `matmult()` multiplies two _or more_ matrices, and can simplify the result 
(with `simplify = TRUE`, default) and/or produce the numeric representation of the product (with `as.numeric = TRUE`, default).

With the exception of `determinant()`, which (because it is a scalar quantity)  returns a character string with a LaTeX expression for the determinant, these operators and functions return `"latexMatrix"` objects, which can be printed, typeset, or used in further computations. Additionally, in many instances a `"latexMatrix"` object can be coerced to a numeric matrix by 
<!-- `as.numeric()` or  -->
`as.double()`. We illustrate these computations in this section.

Consider, first, basic matrix arithmetic. Define some matrices: `A` and `B` are numeric,
but `C` and `D` are symbolic, with elements $c_{ij}$ and $d_{ij}$

```{r matrix-arithmetic}
(A <- latexMatrix(matrix(c(1, -3, 0, 1), 2, 2)))
(B <- latexMatrix(matrix(c(5, 3, -1, 4), 2, 2)))
(C <- latexMatrix(symbol="c", 2, 3))
(D <- latexMatrix(symbol="d", 2, 3))
```

The usual arithmetic operations work for these `"latexMatrix"` objects, and return
LaTeX representations of the result:
```{r matrix-arithmetic2}
A + B
```

Some other examples:
```{r matrix-arithmetic3}
A - B
-A            # unary minus
2*A           # scalar multiplication
C + D         # sum of symbolics
"\\pi" * C    # everything should be multiplied by pi
```

Typesetting the last result produces:
```{r scalar-product-displayed, results='asis', echo=FALSE}
Eqn("\\pi" * C)
```

Some of these operations produce numeric results and so can be coerced to numeric matrices; for example:
```{r coerce-to-numeric}
as.numeric(A + B)
```

Using these tools, it is easy to typeset complete matrix equations, giving multiple
arguments to `Eqn()`:

```{r matrix-arith-equation, results='asis'}
Eqn("\\mathbf{A} + \\mathbf{B} =", A, " + ", B, " = ", A + B)
```

If the elements of a matrix are valid R variable names, then it is also possible to give these elements numeric values, as in
```{r coerce-to-numeric-with-locals}
(M <- latexMatrix(matrix(letters[1:9], 3, 3)))
as.double(-2*M, locals=c(a=1, b=0, c=-2, d=-4, e=7, f=-1, g=0, h=4, i=6))
```

### Matrix products and transpose

The product of two matrices is given by `%*%` for `"latexMatrix"` objects.


```{r product1, results='asis'}
A %*% B
```

To get the numeric result evaluated, use `as.double()` and perform ordinary
matrix multiplication:

```{r product2, results='asis'}
as.double(A) %*% as.double(B) |> latexMatrix()
```

The LaTeX symbol for multiplication is a centered dot, `\\cdot` ($\cdot$), by default.
This can be changed by changing `options(latexMultSymbol)`,
e.g, to use the $\times$ symbol instead, use:

```{r multSymbol, eval=FALSE}
options(latexMultSymbol = "\\times")
```


The transpose, `t()` of `"latexMatrix"` 
objects is similarly straightforward. This directly transposes the matrix, as opposed to superscript notation,
$\mathbf{D}^\top$ or $\mathbf{D}^\prime$ which is implicit.

```{r transpose}
D
t(D)
M %*% t(D)
```
The matrix product in the previous example typesets as
```{r matrix-product-displayed, results='asis', echo=FALSE}
Eqn(M %*% t(D))
```

### Determinants & inverse

The determinant is computed _recursively_ by finding the _cofactors_ of the first row of the matrix,
i.e., $\det(\mathbf{A}_{n \times n}) = \Sigma_j^n \, a_{ij}\: C_{ij}$ where the cofactors
$C_{ij}$ involve the determinants of the $(n-1) \times (n-1)$ minors $\mathbf{A}_{ij}$.
(See the vignette [Evaluation of determinants](http://friendly.github.io/matlib/articles/det-ex2.html) for explanation.)

The method is applicable to a matrix of any order, although beyond an order-3 matrix, the resulting expression gets very long. A couple of examples:
```{r determinant}
A
determinant(A)
M
determinant(M)
```

The determinant of a matrix is a single number. `determinant()` returns the expression that computes it in LaTeX notation, using `\\cdot` to represent $\cdot$ for multiplication.

Typesetting the output from the last command produces
```{r determinant-displayed, results='asis', echo=FALSE}
Eqn(determinant(M))
```

The inverse of a square matrix is computed from its determinant and adjoint matrix; for example:
```{r inverse}
solve(A)
```
Specifying the argument `simplify = TRUE` to `solve()` puts the inverse determinant before the adjoint matrix and returns a latex expression rather than a `"latexMatrix"` object; for example:
```{r simplify-inverse}
solve(M, simplify=TRUE)
```
which typesets as
```{r inverse-displayed, results='asis', echo=FALSE}
Eqn(solve(M, simplify=TRUE))
```
We can also supply values for the elements of the matrix to obtain a numeric inverse:
```{r numeric-inverse}
MASS::fractions(as.double(solve(M), 
                          locals=c(a=1, b=0, c=-2, d=-4, e=7, f=-1, g=0, h=4, i=6)))
MASS::fractions(det(as.double(M, 
                              locals=c(a=1, b=0, c=-2, d=-4, e=7, f=-1, g=0, h=4, i=6))))
```

### Linear hypotheses

As an example of the more general use of these functions, consider the general linear hypothesis used to test hypotheses and contrasts in linear models.
We consider a multivariate regression model 
$\mathbf{Y} = \mathbf{X} \mathbf{B} + \mathbf{E}$ with $q$ regressors
$\mathbf{x}_0, \mathbf{x}_1, \dots, \mathbf{x}_q$ (including the constant
$\mathbf{x}_0$ for the intercept)
and 
$p$ responses,
$\mathbf{y}_1, \mathbf{y}_2, \dots, \mathbf{y}_p$.

<!-- [**Intercept* ignored here. Fix notation or description] -->

Suppose we want to test the hypothesis that a subset of rows (predictors)
and/or columns (responses) simultaneously have null effects.
This can be expressed in the general linear test,
$$
\mathcal{H}_0 : \mathbf{C}_{h \times q} \, \mathbf{B}_{q \times p} = \mathbf{0}_{h \times p} \: ,
$$
where $\mathbf{C}$ is a  full rank $h \le q$ hypothesis matrix of constants, that selects 
subsets or linear combinations (contrasts) of the coefficients in $\mathbf{B}$ to be tested
in a $h$ degree-of-freedom hypothesis.

For example, for a multivariate regression model 
with
three responses $y_1, y_2, y_3$ and three predictors $x_1, x_2, x_3$, the coefficients $\mathbf{B}$ are given by the following `latexMatrix()` expression, where several arguments are used to:
(a) start row indices at zero (`zero.based`);
(b) make the column indices a subscript of $y$ (`prefix.col`);
(c) insert a comma between row/column subscripts.

```{r linhyp-beta, results='asis'}
(B <- latexMatrix('\\beta', ncol = 3, nrow=4, 
                 comma=TRUE, prefix.col = 'y_',
                 zero.based=c(TRUE, FALSE)))
```


We can test the hypothesis
that neither $x_2$ nor $x_3$ contribute at all to the predicting the $y$s in terms of the hypothesis that
the coefficients for the corresponding rows of $\mathbf{B}$ are zero.
To do this, we specify a 2-row $\mathbf{C}$ matrix that
simply selects those rows:

```{r linhyp-C, results='asis'}
(C <- latexMatrix(matrix(c(0, 1, 0, 0,
                           0, 0, 1, 0), nrow=2, byrow=TRUE), 
                 matrix = "bmatrix"))
```

Then, the hypothesis to be tested can be expressed as follows,
using `Eqn()` to wrap a set of LaTeX expressions and calls to `matlib` functions. 

```{r linear-hypotheses, results='asis'}
B0 <- latexMatrix('\\beta', ncol = 3, nrow=2, comma=TRUE, prefix.col = 'y_')
Eqn("\\mathcal{H}_0 : \\mathbf{C} \\mathbf{B} & = ",
    C, B,
    Eqn_newline(), 
    '& =',
    B0,
    "= \\mathbf{0}_{(2 \\times 3)}", 
    align=TRUE)
```

In this example, note that the R objects `C`, `B` and `B0` are the results
of `latexMatrix()` calls, which are character strings containing LaTeX
expressions.

### Partitioned matrices, indexing & binding

Matrix notation sometimes portrays matrices whose elements are themselves
matrices and vectors (rather than scalars) in order to show a higher-level
structure. Such matrices, called **partitioned** or **block** matrices
have similar [arithmetic and algebraic properties](https://www.statlect.com/matrix-algebra/properties-of-block-matrices) to those of ordinary matrices.

For example, the code below represents a $4 \times 4$ matrix
$\mathbf{M}$, which is partitioned in $2 \times 2$ blocks, which are
labeled $\mathbf{M}_{i,j}$.

```{r partition1}
M <- latexMatrix("m", 4, 4)
Mpart <- latexMatrix('\\mathbf{M}', nrow = 2, ncol = 2, comma = TRUE)
Eqn("\\mathbf{M} =", Mpart,
    " =", M)
```

This typesets as:

```{r partition2, results='asis', echo=FALSE}
M <- latexMatrix("m", 4, 4)
Mpart <- latexMatrix('\\mathbf{M}', nrow = 2, ncol = 2, comma = TRUE)
Eqn("\\mathbf{M} =", Mpart,
    " =", M)
```

Just as rows and columns can be selected using `X[rows, cols]` indexing
for ordinary matrices, the same operator can be used for LaTeX matrices,
e.g., `M[rows, cols]`.  The following extracts 4 the sub-matrices of `M`:

```{r partition3}
M11 <- M[1:2, 1:2] |> print()
M12 <- M[1:2, 3:4]
M21 <- M[3:4, 1:2]
M22 <- M[3:4, 3:4]
```

The operations of joining matrices by rows, with `rbind()`, and by columns, with
`cbind()` are also defined for `"latexMatrices"`. This code puts the 4 pieces
of $\mathbf{M}$ back together:

```{r partition4}
rbind(
  cbind(M11, M12),
  cbind(M21, M22)
)
```

And, of course you can also format the sub-matrices together using `Eqn()`:

```{r partition5, results='asis'}
Eqn(M11, M12,
    Eqn_newline(),
    M21, M22,
    align = TRUE)
```

Finally, the `partition()` function alters the print representation of a matrix
using horizontal and  vertical lines separating the sub-matrices.
It does this by re-wrapping the matrix in a LaTeX `\begin{array} ... \end{array}`
environment, using `|` in `{c c | c c}` for the vertical lines and `\hline` for horizontal lines. 
This may be the simplest way to portray partitioned matrices
in writing.

```{r partition6, results='asis'}
partition(M, rows=2, columns=2)
```

Note that `partition()` can show more than one horizontal and vertical partition lines (or no line at all):

```{r partition7, results='asis'}
partition(M, rows=c(1,3), columns=c(1,3))
```

Using this notation, we can illustrate matrix arithmetic with partitioned
matrices. For example, assuming the partitions of matrices $\mathbf{C}$ and $\mathbf{D}$ are of the same size, their sum is just the sum of corresponding
sub-matrices:


```{r partition8, results='asis'}
C <- latexMatrix("\\mathbf{C}", 2, 2)
D <- latexMatrix("\\mathbf{D}", 2, 2)
Eqn("\\mathbf{C} + \\mathbf{D} =",
    C, "+", D, "=", 
    C + D)
```


### Kronecker products

The Kronecker product of two matrices, $\mathbf{A}_{m \times n} \otimes \mathbf{B}_{p \times q}$ is the $mp \times nq$ block matrix consisting of each element $a_{ij}$ multiplied by
$\mathbf{B}$. This has many uses in statistics, among these the nice result [@Bock:75;@Sunwoo1996]
that
the design matrix $\mathbf{X}$ in the linear ANOVA model for factors A, B, C, ...
can be generated as the Kronecker product of their contrast matrices
$\mathbf{C}_A, \mathbf{C}_B, \mathbf{C}_C \dots$, each
preceded by the unit vector $\mathbf{1}$.

$$
\mathbf{X}_{ABC\dots} = 
  [\mathbf{1} \mid \mathbf{C}_A] \;\otimes\;
  [\mathbf{1} \mid  \mathbf{C}_B] \;\otimes\;
  [\mathbf{1} \mid  \mathbf{C}_B] \;\otimes\; \dots
$$

This is implemented in the `%O%` operator and the `kronecker()` function in the package.
For example,

```{r kronecker1, results='asis'}
A <- matrix(1:4, nrow = 2) |> 
  latexMatrix() |> print()
B <- matrix(5:8, nrow = 2) |> 
  latexMatrix() |> print()
kronecker(A, B) |> partition(rows = 2, columns = 2)
```

You can also use `Eqn()` to illustrate the definition of the Kronecker product
more explicitly. In the following, `KAB` is the product in symbolic form;
`as.double()` is used to evaluate the result numerically.


```{r kronecker2, results='asis'}
Bmat <- latexMatrix('\\mathbf{B}', ncol=1, nrow=1)
KABmat <- kronecker(A, Bmat)
KAB <- kronecker(A, B)

Eqn("\\mathbf{A} \\otimes \\mathbf{B} = &",
    KABmat,
    Eqn_newline(), Eqn_vspace("1.5ex"), "= & ",
    KAB |> partition(rows = 2, columns = 2),
    Eqn_newline(), Eqn_vspace("1.5ex"), "= & ",
    latexMatrix(as.double(KAB)) |> partition(rows = 2, columns = 2),
    align = TRUE)
```


## matrix2latex {#matrix2latex}

The `matrix2latex()` function can also generate symbolic equations from numeric or character matrices.
For numeric matrices, it can round the values or show results as fractions.

```{r results='asis'}
A <- matrix(1:12, nrow=3, ncol=4, byrow = TRUE) / 6
matrix2latex(A, fractions = TRUE, brackets = "b") |> Eqn()
```


Say we want to show the matrix $[\mathbf{A} | \mathbf{b}]$ involved in the system of equations
$\mathbf{A} \mathbf{x} = \mathbf{b}$.
Create these as a character matrix and vector:

```{r}
A <- matrix(paste0('a_', 1:9), 3, 3, byrow = TRUE) |> print()
b <- paste0("\\beta_", 1:3) |> print()
```

Then use `matrix2latex()` on `cbind(A,b)` and pipe the result of `matrix2latex()` to `Eqn()`:

```{r results='asis'}
matrix2latex(cbind(A,b)) |> Eqn()
```

All the R tricks for creating and modifying matrices can be used in this way.

## showEqn {#showEqn}

`showEqn()` is designed to show a system of linear equations, $\mathbf{A x} = \mathbf{b}$, but written out as a set of equations individually. With the option `latex = TRUE` it writes these out in LaTeX form.

<!-- `showEqn()` can also write LaTeX, but writes out the equations individually. -->
<!-- However, but the `array` environment needs to be included -->
<!-- inside `$$ ... $$` to be evaluated in a chunk, so this chunk is not evaluated (it causes a LaTeX error). -->

Here, we create a character matrix containing
the elements of a $3 \times 3$ matrix `A`, whose elements are of the form
`a_{ij}` and two character vectors, `b_i` and `x_i`. 
```{r}
A <- matrix(paste0("a_{", outer(1:3, 1:3, FUN  = paste0), "}"), 
            nrow=3) |> print()

b <- paste0("b_", 1:3)
x <- paste0("x", 1:3)
```

`showEqn(..., latex = TRUE)` produces the three equations in a single `\begin{array} ... \begin{array}`
environment.

```{r showEqn0, eval=FALSE}
showEqn(A, b, vars = x, latex=TRUE)
```

If this line was run in an R console, it would produce:

```
\begin{array}{lllllll}
a_{11} \cdot x_1 &+& a_{12} \cdot x_2 &+& a_{13} \cdot x_3  &=&  b_1 \\ 
a_{21} \cdot x_1 &+& a_{22} \cdot x_2 &+& a_{23} \cdot x_3  &=&  b_2 \\ 
a_{31} \cdot x_1 &+& a_{32} \cdot x_2 &+& a_{33} \cdot x_3  &=&  b_3 \\ 
\end{array}
```

Evaluating the above code in an unnumbered LaTeX math environment via `Eqn()` gives the
desired result:

```{r results='asis'}
showEqn(A, b, vars = x, latex=TRUE) |> Eqn()
```

## References
