---
title: "Solving Linear Equations"
author: "Michael Friendly and John Fox"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solving Linear Equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r nomessages, echo = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
options(digits=4)
par(mar=c(5,4,1,1)+.1)
```
```{r setuprgl, echo=FALSE}
library(rgl)
library(knitr)
knit_hooks$set(webgl = hook_webgl)
```


This vignette illustrates the ideas behind solving systems of linear equations of the form $\mathbf{A x = b}$
where 

- $\mathbf{A}$ is an $m \times n$ matrix of coefficients for $m$ equations in $n$ unknowns
- $\mathbf{x}$ is an $n \times 1$ vector unknowns, $x_1, x_2, \dots x_n$
- $\mathbf{b}$ is an $m \times 1$ vector of constants, the "right-hand sides" of the equations

The general conditions for solutions are:

- the equations are *consistent* (solutions exist) if $r( \mathbf{A} | \mathbf{b}) = r( \mathbf{A})$
    - the solution is *unique* if $r( \mathbf{A} | \mathbf{b}) = r( \mathbf{A}) = n$
    - the solution is *underdetermined* if $r( \mathbf{A} | \mathbf{b}) = r( \mathbf{A}) < n$
- the equations are *inconsistent* (no solutions) if $r( \mathbf{A} | \mathbf{b}) > r( \mathbf{A})$

We use `c( R(A), R(cbind(A,b)) )` to show the ranks, and `all.equal( R(A), R(cbind(A,b)) )` to test
for consistency.

```{r}
library(matlib)   # use the package
```
## Equations in two unknowns

Each equation in two unknowns corresponds to a line in 2D space. The equations
have a unique solution if all lines intersect in a point.

### Two consistent equations
```{r}
A <- matrix(c(1, 2, -1, 2), 2, 2)
b <- c(2,1)
showEqn(A, b)
c( R(A), R(cbind(A,b)) )          # show ranks
all.equal( R(A), R(cbind(A,b)) )  # consistent?

```
Plot the equations:
```{r, echo=2}
par(mar=c(4,4,0,0)+.1)
plotEqn(A,b)
```

`Solve()` is a convenience function that shows the solution in a more comprehensible form:
```{r}
Solve(A, b, fractions = TRUE)
```


### Three consistent equations

For three (or more) equations in two unknowns, $r(\mathbf{A}) \le 2$, because  $r(\mathbf{A}) \le \min(m,n)$.
The equations will be consistent
*if* $r(\mathbf{A}) = r(\mathbf{A | b})$. This means that whatever linear relations
exist among the rows of $\mathbf{A}$ are the *same* as those among the elements of $\mathbf{b}$.

Geometrically, this means that all three lines intersect in a point.
```{r}
A <- matrix(c(1,2,3, -1, 2, 1), 3, 2)
b <- c(2,1,3)
showEqn(A, b)
c( R(A), R(cbind(A,b)) )          # show ranks
all.equal( R(A), R(cbind(A,b)) )  # consistent?

Solve(A, b, fractions=TRUE)       # show solution 
```
Plot the equations:
```{r, echo=2}
par(mar=c(4,4,0,0)+.1)
plotEqn(A,b)
```

### Three inconsistent equations
Three equations in two unknowns are *inconsistent* when $r(\mathbf{A}) < r(\mathbf{A | b})$.
```{r}
A <- matrix(c(1,2,3, -1, 2, 1), 3, 2)
b <- c(2,1,6)
showEqn(A, b)
c( R(A), R(cbind(A,b)) )          # show ranks
all.equal( R(A), R(cbind(A,b)) )  # consistent?
```
You can see this in the result of reducing $\mathbf{A} | \mathbf{b}$ to echelon form, where the
last row indicates the inconsistency.
```{r}
echelon(A, b)
```
`Solve()` shows this more explicitly:
```{r}
Solve(A, b, fractions=TRUE)
```


An approximate solution is sometimes available using a generalized inverse.
```{r}
x <- MASS::ginv(A) %*% b
x
```


Plot the equations. You can see that each pair of equations has a solution,
but all three do not have a common, consistent solution.
```{r}
par(mar=c(4,4,0,0)+.1)
plotEqn(A,b, xlim=c(-2, 4))
points(x[1], x[2], pch=15)
```



## Equations in three unknowns

Each equation in three unknowns corresponds to a plane in 3D space. The equations
have a unique solution if all planes intersect in a point.

### Three consistent equations

```{r}
A <- matrix(c(2, 1, -1,
             -3, -1, 2,
             -2,  1, 2), 3, 3, byrow=TRUE)
colnames(A) <- paste0('x', 1:3)
b <- c(8, -11, -3)
showEqn(A, b)
```

Are the equations consistent?
```{r}
c( R(A), R(cbind(A,b)) )          # show ranks
all.equal( R(A), R(cbind(A,b)) )  # consistent?
```
Solve for $\mathbf{x}$.  

```{r}
solve(A, b)
solve(A) %*% b
inv(A) %*% b
```

Another way to see the solution is to reduce $\mathbf{A | b}$ to echelon form. The result is $\mathbf{I | A^{-1}b}$,
with the solution in the last column. 

```{r}
echelon(A, b)
echelon(A, b, verbose=TRUE, fractions=TRUE)
```

Plot them.  `plotEqn3d` uses `rgl` for 3D graphics.  If you rotate the figure, you'll see an orientation
where all three planes intersect at the solution point, $\mathbf{x} = (2, 3, -1)$ 
```{r plotEqn1, webgl=TRUE}
plotEqn3d(A,b, xlim=c(0,4), ylim=c(0,4))
```

### Three inconsistent equations

```{r}
A <- matrix(c(1,  3, 1,
              1, -2, -2,
              2,  1, -1), 3, 3, byrow=TRUE)
colnames(A) <- paste0('x', 1:3)
b <- c(2, 3, 6)
showEqn(A, b)
```
Are the equations consistent? No.
```{r}
c( R(A), R(cbind(A,b)) )          # show ranks
all.equal( R(A), R(cbind(A,b)) )  # consistent?
```
